---
title: "SARS-CoV-2 dataset: Running CARNIVAL in different Conditions"
author: "Alberto Valdeolivas: alberto.valdeolivas@bioquant.uni-heidelberg.de; Date: "
date: "14/04/2020"
output: pdf_document
---

```{r setup, include=FALSE, warning=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### License Info

This program is free software: you can redistribute it and/or modify it under 
the terms of the GNU General Public License as published by the Free Software 
Foundation, either version 3 of the License, or (at your option) any later 
version.

This program is distributed in the hope that it will be useful, but WITHOUT ANY 
WARRANTY; without even the implied warranty of MERCHANTABILITY or FITNESS FOR 
A PARTICULAR PURPOSE. See the GNU General Public License for more details.

Please check http://www.gnu.org/licenses/.

## Introduction

The present script deal with the RNAseq data from the study *"SARS-CoV-2 launches* 
*a unique transcriptional signature from in vitro, ex vivo, and in vivo systems"* 

<https://www.biorxiv.org/content/10.1101/2020.03.24.004655v1>

<https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE147507>

It uses the pathway activity scores from **Progeny** and the TFs activity scores 
from **Dorothea** obtained in the previous script to run **CARNIVAL**. In 
addition, CARNIVAL requires a prior knoledge network that would be obtained from 
**Omnipath** via OmnipathR. CARNIVAL is launched using different assumptions 
for the four condiitons under study:

+ Human lung epithelial cells (NHBE): mock treated vs infected with SARS-CoV-2. 

+ A549 alveolar cancer cell line: mock treated vs infected with SARS-CoV-2. 

+ A549 cell line does not express ACE2, the receptor used by SARS-CoV-2 to 
penetrate into human cells. Therefore A549 were also transduced with ACE2 and 
then mock treated or infected with SARS-CoV-2

+ Calu-3 human lung epithelial cancer cell line: mock treated vs 
infected with SARS-CoV-2.   

## Getting Started

We first load the required libraries. 

```{r}
# install_github("saezlab/progeny")
# install_github("saezlab/dorothea")
#install_github('COVID-19-Causal-Reasoning/CARNIVAL', ref="add_gurobi", force=TRUE)
library(argparser)
library(readr)
# 
p <- arg_parser("Run CARNIVAL on DOROTHEA and PROGENy output")
p <- add_argument(p, "pathway_activities_file",  help="Pathway activities  (from Progeny)")
p <- add_argument(p, "tf_activities_file", help="Transcription Factor Activities (From Dorothea)")
p <- add_argument(p, "--perturbation_file", help="Perturbation File", default="NoInput")
p <- add_argument(p, "--variable", help="Intervention (e.g. VN1203)")
p <- add_argument(p, "--baseline", help="The thing to be compared against. E.g. Mock")
p <- add_argument(p, "--constant", help="The things that are held constant for both variable and baseline. For example timepoint (7h) or strain (VN1203")
p <- add_argument(p, "--outdir", help="output directory")
p <- add_argument(p, "--timelimit", help="Number of seconds before timeout. Default 1200", default=1200)
p <- add_argument(p, "--mipgap", help="MIP Gap. default: 3%", default=0.03)
p <- add_argument(p, "--threads", help="number of threads", default=8)
p <- add_argument(p, "--solver", "cplex gurobi lpsolve or cbc", default="cplex")
p <- add_argument(p, "--solver_path", "location of solver binary", default="/Applications/CPLEX_Studio_Community129/cplex/bin/x86-64_osx/cplex")
argv <- parse_args(p) #, c( "IntermediateFiles/VN1203vsMock_for_24h_pathways_activity_score_inputCarnival.csv",
                      #  "IntermediateFiles/VN1203vsMock_for_24h_tf_activities_stat_inputCarnival.csv",
                      #  "--variable", "VN1203", 
                      #  "--perturbation_file", "kegg_influenza_ns1.txt",
                      #  "--constant", "24hr",
                      #  "--timelimit", "150",
                      #  "--mipgap", "0.40",
                      #  "--outdir", "ResultsCARNIVAL_VN1203vsMock_for_24hr"))



Sys.setenv(CPLEX_STUDIO_KEY='api_cos_0c913dc4-503c-4334-9b4e-ae4c7d13158b')
Sys.setenv(CPLEX_STUDIO_DIR1210 = '/Applications/CPLEX_Studio_Community129' )
carnival_dir <-"~/Projects/CausalInference/Pathogenesis/OmniPath/CARNIVAL/R"
for(filename in list.files(carnival_dir)){
  source(paste( carnival_dir, filename, sep="/"))}

#library(CARNIVAL)
```

```{r}
library(rlist)
library(devtools)
library(tidyverse)
library(OmnipathR)
library(dplyr)
library(tibble)
library(openxlsx)

## We also define a function to format the CARNIVAL output to cytoscape
OutputCyto <- function(CarnivalResults, outputFile) {
    CarnivalNetwork <- 
        as.data.frame(CarnivalResults$weightedSIF, stringsAsFactors = FALSE) 
    
    CarnivalNetworkNodes <- 
        unique(c(CarnivalNetwork$Node1,CarnivalNetwork$Node2))
    
    CarnivalAttributes <- CarnivalResults$nodesAttributes %>% 
        as.data.frame() %>%
        dplyr::filter(Node %in% CarnivalNetworkNodes) %>%
        dplyr::mutate(NodeType = as.character(NodeType)) %>%
        dplyr::mutate(NodeType=if_else(NodeType =="", "I", NodeType))
            
    nameOutputNetwork <- paste0(outputFile, "Network.sif")
    nameOutputAttributes <-  paste0(outputFile, "Attributes.txt")    
    
    write.table(CarnivalNetwork[,c(1,2,3)], file = nameOutputNetwork,
        quote = FALSE, row.names = FALSE, col.names = FALSE, sep = " ")
    
    write.table(CarnivalAttributes, file = nameOutputAttributes,
        quote = FALSE, row.names = FALSE, col.names = TRUE, sep = "\t")
}
```


## Prior Knowledge Network from Omnipath

**CARNIVAL** aims at finding the most likely signaling paths in a given 
situation. In order to do so, it requires a prior knolwdge network that 
we extract from Omnipath. 

```{r, message=FALSE}
## the OmniPath PPI interaction network
ia_omnipath <- import_Omnipath_Interactions() %>% as_tibble()

## We also download the other datasets containing interactions
ia_ligrec <- import_LigrecExtra_Interactions() %>% as_tibble()
ia_pwextra <- import_PathwayExtra_Interactions() %>% as_tibble()
ia_kinaseextra <- import_KinaseExtra_Interactions() %>% as_tibble()

## We bind the datasets
interactions <- as_tibble(
    bind_rows(
        ia_omnipath %>% mutate(type = 'ppi'),
        ia_pwextra %>% mutate(type = 'ppi'),
        ia_kinaseextra %>% mutate(type = 'ppi'),
        ia_ligrec %>% mutate(type = 'ppi')))

## I am going to keep only directed interactions (consensus_direction) and 
## signed interactions (consensus_stimulation/consensus_inhibition)
## We transform to the format needed by CARNIVAL. We just keep signed and 
## directed interactions 
SignedDirectedInteractions <- 
    dplyr::filter(interactions, consensus_direction==1) %>%
    filter(consensus_stimulation == 1 | consensus_inhibition == 1)

NetworkCarnival_df <- bind_rows(
  (SignedDirectedInteractions %>%
  filter(consensus_stimulation == 1 & consensus_inhibition == 0) %>%
  transmute(source_genesymbol, interaction = 1, target_genesymbol)),   
  (SignedDirectedInteractions %>%
     filter(consensus_stimulation == 0 & consensus_inhibition == 1) %>%
     transmute(source_genesymbol, interaction = -1, target_genesymbol))) %>%  
  distinct() 

## We transform the network to an igraph object to simplify
NetworkCarnival_igraph <- 
    graph_from_data_frame(NetworkCarnival_df[c(1,3,2)], directed = TRUE) %>% 
    igraph::simplify(remove.multiple = TRUE, remove.loops = TRUE, 
        edge.attr.comb = "first")

## We transform back to the format required by CARNIVAL
NetworkCarnival_df <- igraph::as_data_frame(NetworkCarnival_igraph) %>%
    dplyr::select(from, interaction, to) %>%  
    distinct() 

## We have to be careful with the gene names with a "-". CPLEX gets crazy. 
NetworkCarnival_df$from <- gsub("-","_", NetworkCarnival_df$from)
NetworkCarnival_df$to <- gsub("-","_", NetworkCarnival_df$to)

AllNodesNetwork <- unique(c(NetworkCarnival_df$from, NetworkCarnival_df$to))
```
We also read pathway and TFs activity scores from the previous script. 

```{r}
pathways_activity_score_inputCarnival <- as.matrix(read.csv(argv$pathway_activities_file, row.names=1))
tf_activities_stat <- as.matrix(read.csv(argv$tf_activities_file, row.names=1))
```


## Running CARNIVAL

We now run CARNIVAL for both cell lines, NHBE and A549, under different 
assumptions

### Selecting the most active TFs

In order to obtain a network that we can visualize and interpret, we select 
the top 50 most active/inactive TFs present in our prior knowledge network.

```{r, eval=TRUE, echo=TRUE}
### VN1203 vs Mock
tf_activities_stat_top50 <- tf_activities_stat %>% 
  as.data.frame() %>% 
  rownames_to_column(var = "GeneID") %>%
  dplyr::filter(GeneID %in% AllNodesNetwork) %>%
  dplyr::arrange(desc(abs(stat))) %>%
  dplyr::top_n(50, wt = abs(stat)) %>%
  column_to_rownames(var = "GeneID") %>%
  t()
write.csv(colnames(tf_activities_stat_top50), 
   paste0(argv$outdir, "/Top50_tf_activities_", argv$variable, "vs", argv$baseline, "_for_", argv$constant))


```


### CARNIVAL with perturbations file


```{r}
if ( argv$perturbation_file == "NoInput"){
  HostVirus_perturbation = NULL
  NetworkCarnivalHostVirus_df <- NetworkCarnival_df
} else {
 
  perturbation <- read.csv(argv$perturbation_file, sep="\t")

  NetworkCarnivalHostVirus_df <- 
    bind_rows(NetworkCarnival_df, perturbation) %>% 
    dplyr::distinct()

## We have to be careful with the gene names with a "-". CPLEX gets crazy. 
  NetworkCarnivalHostVirus_df$from <- 
    gsub("-","_", NetworkCarnivalHostVirus_df$from)
  NetworkCarnivalHostVirus_df$to <- 
   gsub("-","_", NetworkCarnivalHostVirus_df$to)

  Virusproteins <- unique(perturbation$from)
  Virusproteins <-  gsub("-","_", Virusproteins)
  HostVirus_perturbation <- 
   data.frame(Virusproteins = rep(1,length(Virusproteins)))
  rownames(HostVirus_perturbation) <- Virusproteins
  HostVirus_perturbation <- as.data.frame(t(as.matrix(HostVirus_perturbation)))
} 
  
```
#### VN1203 cell line

First for the VN1203 line:

```{r, message=FALSE, eval=TRUE}
CarnivalResults <-runCARNIVAL(
    solverPath=argv$solver_path,
      #"/Applications/CPLEX_Studio_Community129/cplex/bin/x86-64_osx/cplex",
   # solverPath="/usr/local/bin/cbc",
   #solverPath="/Library/gurobi900/mac64/bin/gurobi_cl",
    netObj= NetworkCarnivalHostVirus_df,
    measObj=as.data.frame(tf_activities_stat_top50),
    inputObj = HostVirus_perturbation,
    mipGAP = argv$mipgap,
    threads = argv$threads,
    # DOTfig=TRUE, 
    dir_name=argv$outdir,
    weightObj=t(pathways_activity_score_inputCarnival),
    # nodeID = 'gene',
    timelimit = argv$timelimit,
    solver = argv$solver)
```

```{r}
if (argv$perturbation_file == "NoInput") {
  saveRDS(CarnivalResults, file = paste0(argv$outdir, "/", argv$variable, "vs", argv$baseline, "_for_", argv$constant, "_noPerturbation.rds"))
  OutputCyto(CarnivalResults, 
  outputFile= paste0(argv$outdir, "/", argv$variable, "vs", argv$baseline, "_for_", argv$constant, "_noPerturbation"))
} else {
   saveRDS(CarnivalResults, file = paste0(argv$outdir, "/", argv$variable, "vs", argv$baseline, "_for_", argv$constant, "_HostVirusPerturbation.rds"))
   OutputCyto(CarnivalResults, 
  outputFile= paste0(argv$outdir, "/", argv$variable, "vs", argv$baseline, "_for_", argv$constant, "_HostVirusPerturbation"))
}

```